---
title: "Collaborator Analysis of Lobster Trap Counts"
author: "Tymesha Bovell"
format: html
---

# Introduction:

California spiny lobser (Panulirus interruptus)are important predators in southern California giant kelp forests. They are studied along the coast of the Santa Barbara Channel to understand their patterns in abundance, size and fishing pressure as these variables affect their role in kelp forest community dynamics. Fishing pressure is estimated from using trap counts collected during the lobster catch season (October-March).

In this project a dataset from 2012 from inside and outside Marine Protected Areas (MPAs) along the Santa Barbara Coast was collected to investigate differences between protected and non-protected areas. This analysis focuses on trap counts as a proxy for fishing pressure which may explain differences in lobster abundance and size observed in other parts of the dataset.

In this analysis, I examine how trap counts vary across sites and years to better understand patterns in fishing pressure and how they may differ across locations.

```{r}
# Loading libraries
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(here)

# Read in trap dataset and replace missing values
lobster_traps <- read_csv(here::here("data", "Lobster_Trap_Counts_All_Years_20210519.csv")) %>%
  mutate(TRAPS = na_if(TRAPS, -99999))


```

# Data Exploration and Cleaning

The lobster trap buoy count dataset contains observations of commercial trap floats used as a proxy for fishing pressure along the mainland coast of the Santa Barbara Channel. Fishing pressure is estimated by counting the number of lobster traps present at each site during the fishing season from October through March, with surveys conducted every two to four weeks. Two of the study reefs, Naples (NAPL) and Isla Vista (IVEE), became marine protected areas (MPAs) in January 2012, which makes this dataset useful for comparing fishing pressure between protected and non-protected locations.

The trap dataset includes variables such as fishing season, year, month, date, site, and trap count (`TRAPS`). During cleaning, the `TRAPS` column was checked for placeholder missing values, and `-99999` was recoded as `NA` so that missing observations would not distort summaries or visualizations.

# Data Wrangling

To support interpretation of fishing pressure, trap counts were summarized by site and year to examine temporal changes across locations. In addition, observations from 2019 to 2021 were classified into "high" and "low" fishing pressure categories using a threshold of 8 traps, following the class exercise. These categories were then summarized by site to compare how frequently higher fishing pressure occurred across the study area. \# Data Visualization

## Figure 4: Distribution of Trap Counts by Year

It shows how lobster trap counts were distributed across years in the monitoring record.

```{r}
ggplot(data = lobster_traps, aes(x = TRAPS)) +
  geom_histogram(
    bins = 30, 
    fill = "steelblue", 
    color = "white"
  ) +
  
  facet_wrap(~ YEAR, ncol = 4) +
  
  labs(
    title = "Distribution of Lobster Trap Counts by Year",
    x = "Number of Traps",
    y = "Frequency (Observation Counts)"
  ) +
  
  scale_x_continuous(
    breaks = seq(0, 40, by = 10)
  ) +
  
  theme_light() +
  
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    strip.text = element_text(size = 10, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

## Figure 5: Temporal Trends in Trap Counts

It tracks total lobster trap counts at each site across the study period.

```{r}
# Summarize trap counts by site and year
lobster_traps_summarize <- lobster_traps %>%
  group_by(SITE, YEAR) %>%
  summarise(TOTAL_TRAPS = sum(TRAPS, na.rm = TRUE), .groups = "drop")

# Line and point plot
ggplot(data = lobster_traps_summarize, 
       aes(x = YEAR, y = TOTAL_TRAPS, color = SITE)) +
  geom_line() +
  geom_point() +
  
  # Add title + labels
  labs(
    title = "Total Lobster Trap Counts by Site Over Time",
    x = "Year",
    y = "Total Trap Counts",
    color = "Site"
  ) +
  
  # Clean theme + center title
  theme_light() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```
## Figure 6: Fishing Pressure Categories During 2019–2021

It compares how often high and low fishing pressure occurred at each site from 2019 to 2021.

```{r}
# Create fishing pressure categories
lobster_traps_fishing_pressure <- lobster_traps %>%
  filter(YEAR %in% c(2019, 2020, 2021)) %>%
  mutate(FISHING_PRESSURE = if_else(TRAPS >= 8, "high", "low")) %>%
  group_by(SITE, FISHING_PRESSURE) %>%
  summarise(COUNT = n(), .groups = "drop") %>%
  drop_na()


# Create a styled plot object for fishing pressure by site
trap_pressure_plot <- ggplot(
  data = lobster_traps_fishing_pressure,
  aes(x = SITE, y = COUNT, fill = FISHING_PRESSURE)
) +
  geom_col(position = "dodge") +              # side-by-side bars
  coord_flip() +                              # flip for easier reading
  labs(
    title = "Fishing Pressure by Site (2019–2021)",
    x = "Site",
    y = "Count",
    fill = "Fishing Pressure"
  ) +
  theme_light() +                             # cleaner overall theme
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),   # center + bold title
    axis.title.y = element_text(face = "bold"),              # bold y-axis title
    axis.title.x = element_text(face = "bold"),              # bold x-axis title
    text = element_text(size = 12)                           # readability
  )

# Display plot
trap_pressure_plot
```
